---
title             : "Factor Loading Recovery for Smoothed Non-positive Definite Correlation Matrices"
shorttitle        : "Loading Recovery for Smoothed Matrices"

author: 
  - name          : "Justin D. Kracht"
    affiliation   : "1"
    corresponding : yes
    address       : "Postal address"
    email         : "krach018@umn.edu"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, University of Minnesota"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  
  One or two sentences to put the results into a more **general context**.
  
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

## Matrix Smoothing Algorithms

### Higham Alternating Projections Algorithm (APA; 2002)

### Bentler-Yuan Algorithm (BY; 2011)

### Knol-Berger Algorithm (KB; 1991)

## Factor Estimation Methods

### Principal Axes Factor Analysis

### Least-Squares Factor Analysis

### Maximum-Likelihood Factor Analysis

# Methods



# Results

# Discussion

\newpage

# References

```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
